### 2024-06-02 2024-5-13/sections/heuristic.tex
section 7.3

It turns out that this model generation process is inefficient, mainly due to the fact that the information about the occurrences of vectors $\myvec{v}$ in $\cB_i$ in the searching process is too coarse. In the sequel, we would like to refine this process by using the LIA formulas to obtain the information about the number of occurrences of transitions, instead of just the integer vectors, and utilize this information to guide the searching process. Moreover, some additional tricks are introduced to prune the searching space further.

\zhilin{stopped here}

In the section \ref{sec:algorithm}, we present an algorithm to solve the $\cefadec$ problem. The algorithm computes existential LIA formulas and then uses SMT solver to solve it. However, this procedure only anwser the question whether the language of the regex is empty or not. If the language is not empty, we need to find a word in the language in practice. In our preivous paper \cite{Denghang2023}, we do not explicitly present the algorithm to find a word in the language because of the space limitation. We have used a naive approach which randomly searching the CEFAs by the depth-first search, based on the regesiters values obtained from solving the existential LIA formula. However, this approach is not efficient when the regexes is non-regesiter-representable. Suppose that there are $n$ registers whose values are $r_1,\cdots, r_n$ respectively in the CEFA, we may visit each transition $O(r_1*r_2\cdots *r_n)$ times during the search. To relieve this problem, we present a heuristic to find a word in the language as follows.

@@

  In Algorithm~\ref{alg:isReachable}, all transitions reachable from the state $q'$ is computed and saved to $qreach$. If $qreach$ covers all transitions with positive times in the input map $tmap$, then we say that $tmap$ is reachable from $q'$ and the procedure \textsc{isReachable} returns true, otherwise it returns false.

>>>>

It turns out that this model generation process is inefficient, mainly due to the fact that the information about the occurrences of vectors $\myvec{v}$ in $\cB_i$ in the searching process is too coarse. In the sequel, we would like to refine this process by using the LIA formulas to obtain the information about the number of occurrences of transitions, instead of just the occurrences of integer vectors, and utilize this information to guide the search process. 
Specifically, for each $i \in [n]$ and each transition $(q, a, q', \myvec{v})$ in $\cB_i$, we introduce an integer variable $\anivarz_{i, (q, a, q', \myvec{v})}$ to denote the number of occurrences of the transition, and solve the satisfiability of the following formula,  
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right]\ \wedge \\
\ \ \bigwedge \limits_{i \in [n]} \bigwedge \limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} = \sum \limits_{(q, a, q', \myvec{v}) \mbox{ occurs in } \cB_i} \anivarz_{i, (q, a, q', \myvec{v})}.
\end{array}
\]
If an SMT solver reports that the aforementioned formula is satisfiable, then the solver can also returns a model, which assigns to each variable $\anivarz_{i, (q, a, q', \myvec{v})}$ a natural number. 
%For each $i \in [n]$ and each transition $(q, a, q', \myvec{v})$ in $\cB_i$, let us use $nt_{i, (q, a, q', \myvec{v})}$ to denote the value of $\anivarz_{i, (q, a, q', \myvec{v})}$ in the model.
Then we utilize these values to generate a model (an assignment of strings to string variables) of $\varphi$. 
Let $i \in [n]$. We use a function $\lambda_i$ to record the number of occurrences of the transitions in $\cB_i$. During the search process for $\cB_i$, each time when a transition $(q, a, q', \myvec{v})$ is traversed, $\lambda_i$ is updated by decrementing $\lambda_i((q, a, q', \myvec{v}))$, moreover, $a$ is also recorded. Initially, for each $i \in [n]$ and each transition $(q, a, q', \myvec{v})$ in $\cB_i$, $\lambda_i((q, a, q', \myvec{v}))$ is set to be the value of $\anivarz_{i, (q, a, q', \myvec{v})}$ in the model. When a final state of $\cB_i$ is reached and $\lambda_i((q, a, q', \myvec{v}))$ becomes zero for each transition $(q, a, q', \myvec{v})$ in $\cB_i$, then a desired string is generated for the string variable $x_i$.

Moreover, propose a method to prune the search space further. Let $i \in [n]$. Suppose that during the search process for $\cB_i$, let $q$ be the current state of $\cB_i$ and $\lambda_i$ record the remaining occurrences of transitions of $\cB_i$. We compute the set of transitions of $\cB_i$ that are reachable from the state $q$. Let $\cT_q$ denote this set. If $\{(q, a, q', \myvec{c}) \mid \lambda_i((q, a, q', \myvec{c})) > 0\} \not \subseteq \cT_q$, then we know that it is impossible to continue the search process so that a string accepted by $\cB_i$ will be found and meanwhile the constraint on the number of transitions enforced by $\lambda_i$ is satisfied. Therefore, in this case, the search process for $\cB_i$ in this direction halts and the search process is backtracked.

%\zhilin{stopped here}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hide{
In the section \ref{sec:algorithm}, we present an algorithm to solve the $\cefadec$ problem. The algorithm computes existential LIA formulas and then uses SMT solver to solve it. However, this procedure only anwser the question whether the language of the regex is empty or not. If the language is not empty, we need to find a word in the language in practice. In our preivous paper \cite{Denghang2023}, we do not explicitly present the algorithm to find a word in the language because of the space limitation. We have used a naive approach which randomly searching the CEFAs by the depth-first search, based on the regesiters values obtained from solving the existential LIA formula. However, this approach is not efficient when the regexes is non-regesiter-representable. Suppose that there are $n$ registers whose values are $r_1,\cdots, r_n$ respectively in the CEFA, we may visit each transition $O(r_1*r_2\cdots *r_n)$ times during the search. To relieve this problem, we present a heuristic to find a word in the language as follows.

@@

  In Algorithm~\ref{alg:isReachable}, all transitions reachable from the state $q'$ is computed and saved to $qreach$. If $qreach$ covers all transitions with positive times in the input map $tmap$, then we say that $tmap$ is reachable from $q'$ and the procedure \textsc{isReachable} returns true, otherwise it returns false.  
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-06-02 2024-5-13/sections/implementation.tex
section 7.3

We implemented the algorithm in Section~\ref{sec:algorithm} and Section~\ref{sec:heuristic} on top of OSTRICH, resulted to a string solver called $\ostrichrecl$. 
%
In this section, we evaluate the performance of $\ostrichrecl$ on three benchmark suites, that is, RegCoL, AutomatArk, and ComplexCounting. In the sequel, we first describe the three benchmark suites as well as the experiment setup. Then we present the experiment results. We evaluate the performance and correctness of our solver against the state-of-the-art string constraint solvers, including CVC5 \cite{cvc5}, Z3seq \cite{z3seq}, Z3str3 \cite{Z3-str3}, Z3str3RE \cite{BD+23}, and OSTRICH \cite{CHL+19}. We also compare $\ostrichrecl$ with several variants of it, to justify the technical choices and heuristics we take in the Section~\ref{subsec:cefadec} and Section~\ref{sec:heuristic}.

@@

\subsection{Benchmark suites and experiment setup}\label{sec:bench}

Our experiments utilize two benchmark suites, namely, \emph{RegCoL} and \emph{AutomatArk}. Other industrial benchmark suites are not utilized because they contain no counting operators. There are 49,843 instances in total, where 48,843 of them are collected from the real world and 1,000 of them are generated by fuzzing tool. All benchmark instances are in the SMTLIB2 format.

@@

\medskip
\noindent
\emph{ComplexCounting benchmark suite.} This benchmark suite is generated by string fuzzing test tool \textsc{stringfuzz} \cite{stringfuzz} extended with the counting operator. We generate 500 instances of the nesting of counting operators containing the RECL constraint $x\in e_{nest}\wedge x\in e_{sani}\wedge |x|>50$ in each of them, where $e_{nest}\equiv (e_0^{\{m_0,n_0\}}e_1^{\{m_1,n_1\}}e_2^{\{m_2,n_2\}}e_3^{\{m_3,n_3\}})^{\{m_4,n_4\}}$ is a regex of the nesting of counting operators and $e_{sani} \equiv \overline{\Sigma^*(<+ >+'+''+\&)\Sigma^*}$ is as RegCoL benchmark suite. Note that $0\leq m_i\leq n_i\leq 20$ for $i\in [4]$ and $e_i$ containing no counting operators for $i\in [3]$ is randomly generated.  We also generate 500 instances of the nesting of counting operators and a complement operator containing the RECL constraint $x\in e_{nest}\wedge x\in e_{comp} \wedge x\in e_{sani} \wedge |x|>50$ in each of them, where $e_{comp}\equiv \overline{e_0^{\{m_0,n_0\}}e_1^{\{m_1,n_1\}}e_2^{\{m_2,n_2\}}e_3^{\{m_3,n_3\}}}$ is a regex of the nesting of counting operators and a complement operator, and $e_{nest}$ and $e_{sani}$ are as above. Note that $20\leq m_i\leq n_i\leq 1000$ and $e_i$  containing no counting operators for $i\in [3]$ is randomly generated.

@@

We evaluate the performance of $\ostrichrecl$ against the state-of-the-art string constraint solvers, including CVC5
\cite{cvc5}, Z3seq \cite{z3seq}, Z3str3
\cite{Z3-str3}, Z3str3RE \cite{BD+23}, and OSTRICH
\cite{CHL+19}, on RegCoL, AutomatArk and ComplexCounting benchmark suites.
The experiment results can be found in Figure~\ref{fig:table_overall_eval}. Note that we take the results of CVC5 as the ground truth\footnote{Initially,  we used the majority vote of the results of the solvers as the ground truth. Nevertheless, on some problem instances, all the results of the three solvers in the Z3 family are wrong (after manual inspection), thus failing this approach on these instances.}, and the results different from the ground truth are classified as \emph{error}. We can see that $\ostrichrecl$ solves almost all instances, except 182 of them, that is, it solves 49,638 instances correctly. The number is 4,547/1,917/13,618/2,764/3,134 more than the number of instances solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively.
%
Moreover, $\ostrichrecl$ is the second fastest solver, whose average time on each instance is close to the fastest solver Z3str3RE (2.88s versus 1.99s).

\begin{table}
  \centering\vskip 0pt
    \import{tables}{table_regcol.tex}
    \caption{Overall performance evaluation}
  \label{fig:table_overall_eval}
\end{table}
%To answer \textbf{Q1}, \textbf{$\ostrichrecl$ is the second fastest solver on these benchmark suites, and it solves the most number of instances}.

@@

\subsection{Empirical justification of the technical choices and optimizations}\label{subsec:size_reduction_eval}
We also do experiments to evaluate the effectiveness of the automata size-reduction techniques for register-representable regexes, that is, Step 2 of Section~\ref{subsec:cefadec}. We practically answer why we choose LIA formula to solve the nonemptiness problem instead of using the nuXmv model checker like \cite{atva2020}. Finally, we carry out experiment on the ComplexCounting benchmarks to justify the optimizations of Section~\ref{sec:heuristic} for non-register-representable regexes.
\subsubsection{Empirical justification of the technical choices}
%We also do experiments to evaluate the effectiveness of the automata size-reduction techniques, that is, Step 2 in Section~\ref{subsec:cefadec}. 
%Let us use $\ostrichrecl_{\rm -SIMP}$ to denote $\ostrichrecl$ with the size-reduction techniques removed. 
We compare $\ostrichrecl$ with its variant  ${\rm +NUXMV}$. ${\rm +NUXMV}$  is obtained from $\ostrichrecl$ by using the nuXmv model checker to solve the nonemptiness of $\cefadec$.

@@

Since the non-register-representable regexes collected from real world are few (5\% of RegCoL and AutomatArk benchmark suites) and simple, there seems no much difference between the performance of $\ostrichrecl$ and its variants ${\rm -NEST}$, ${\rm -COMP}$, ${\rm -FIND}$ in the overall experiment results. Therefore, we consider the ComplexCounting benchmark suite, which contains more complex non-register-representable regexes, to justify the optimizations for non-register-representable regexes. The ComplexCounting benchmark experiment results is shown in Table~\ref{tab:results_heuristic_complex}. We can see that $\ostrichrecl$ solves 82/339/326 more instances and is 1.02/1.92/1.61 times faster than ${\rm -NEST}$/${\rm -COMP}$/${\rm -FIND}$ respectively. From the results, we can conclude that all the optimizations sharply improve the performance of $\ostrichrecl$ on the non-register-representable regexes.

\begin{table}
  \import{tables}{table_heuristics_complex.tex}
  \caption{Empirical justification of the optimizations on ComplexCounting benchmarks}\label{tab:results_heuristic_complex}
\end{table}

>>>>

We implemented the procedures and optimizations in Section~\ref{sec:algorithm}-\ref{sec-opt-sol-gen} on top of OSTRICH, resulted to a string solver called $\ostrichrecl$. 
%
In this section, we evaluate the performance of $\ostrichrecl$ on three benchmark suites, that is, RegCoL, AutomatArk, and NestC. In the sequel, we first describe the three benchmark suites as well as the experiment setup. Then we present the experiment results. We evaluate the performance and correctness of our solver against the state-of-the-art string constraint solvers, including CVC5 \cite{cvc5}, Z3seq \cite{z3seq}, Z3str3 \cite{Z3-str3}, Z3str3RE \cite{BD+23}, and OSTRICH \cite{CHL+19}. We also compare $\ostrichrecl$ with several variants of it, to justify the technical choices made in Section~\ref{sec:algorithm} and evaluate the effectiveness of the optimizations in Section~\ref{sec:heuristic}-\ref{sec-opt-sol-gen}.

@@

\subsection{Benchmark suites and the experiment setup}\label{sec:bench}

Our experiments utilize three benchmark suites, namely, \emph{RegCoL}, \emph{AutomatArk}, and \emph{NestC}. 
%Other benchmark suites are not utilized because they contain no counting operators. 
There are 49,843 instances in total.
%, where 48,843 of them are collected from the real world and 1,000 of them are generated by the string fuzzing tool. 
All benchmark instances are in the SMTLIB2 format. In the sequel, we give more details of these three benchmark suites.

@@

\medskip
\noindent
\emph{NestC benchmark suite.} 
\zhilin{To be cleaned}
This benchmark suite is generated by string fuzzing test tool \textsc{stringfuzz} \cite{stringfuzz} extended with the counting operator. We generate 500 instances of the nesting of counting operators containing the RECL constraint $x\in e_{nest}\wedge x\in e_{sani}\wedge |x|>50$ in each of them, where $e_{nest}\equiv (e_0^{\{m_0,n_0\}}e_1^{\{m_1,n_1\}}e_2^{\{m_2,n_2\}}e_3^{\{m_3,n_3\}})^{\{m_4,n_4\}}$ is a regex of the nesting of counting operators and $e_{sani} \equiv \overline{\Sigma^*(<+ >+'+''+\&)\Sigma^*}$ is as RegCoL benchmark suite. Note that $0\leq m_i\leq n_i\leq 20$ for $i\in [4]$ and $e_i$ containing no counting operators for $i\in [3]$ is randomly generated.  We also generate 500 instances of the nesting of counting operators and a complement operator containing the RECL constraint $x\in e_{nest}\wedge x\in e_{comp} \wedge x\in e_{sani} \wedge |x|>50$ in each of them, where $e_{comp}\equiv \overline{e_0^{\{m_0,n_0\}}e_1^{\{m_1,n_1\}}e_2^{\{m_2,n_2\}}e_3^{\{m_3,n_3\}}}$ is a regex of the nesting of counting operators and a complement operator, and $e_{nest}$ and $e_{sani}$ are as above. Note that $20\leq m_i\leq n_i\leq 1000$ and $e_i$  containing no counting operators for $i\in [3]$ is randomly generated.

@@

We evaluate the performance of $\ostrichrecl$ against the state-of-the-art string constraint solvers, including CVC5
\cite{cvc5}, Z3seq \cite{z3seq}, Z3str3
\cite{Z3-str3}, Z3str3RE \cite{BD+23}, and OSTRICH
\cite{CHL+19}, on RegCoL, AutomatArk and NestC benchmark suites.
The experiment results can be found in Table~\ref{tab:table_overall_eval}. Note that we take the results of CVC5 as the ground truth\footnote{Initially,  we used the majority vote of the results of the solvers as the ground truth. Nevertheless, on some problem instances, all the results of the three solvers in the Z3 family are wrong (after manual inspection), thus failing this approach on these instances.}, and the results different from the ground truth are classified as \emph{error}. We can see that $\ostrichrecl$ solves almost all instances, except 182 of them, that is, it solves 49,638 instances correctly. The number is 4,547/1,917/13,618/2,764/3,134 more than the number of instances solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively.
%
Moreover, $\ostrichrecl$ is the second fastest solver, whose average time on each instance is close to the fastest solver Z3str3RE (2.88s versus 1.99s).

\begin{table}
  \centering\vskip 0pt
    \import{tables}{table_regcol.tex}
    \caption{Overall performance evaluation}
  \label{tab:table_overall_eval}
\end{table}
%To answer \textbf{Q1}, \textbf{$\ostrichrecl$ is the second fastest solver on these benchmark suites, and it solves the most number of instances}.

@@

\subsection{Empirical justification of the technical choices and optimizations}\label{subsec:size_reduction_eval}
We also do experiments to evaluate the effectiveness of the automata size-reduction techniques for register-representable regexes, that is, Step 2 of Section~\ref{subsec:cefadec}. We practically answer why we choose LIA formula to solve the nonemptiness problem instead of using the nuXmv model checker like \cite{atva2020}. Finally, we carry out experiment on the NestC benchmarks to justify the optimizations of Section~\ref{sec:heuristic} for non-register-representable regexes.
\subsubsection{Empirical justification of the technical choices}
%We also do experiments to evaluate the effectiveness of the automata size-reduction techniques, that is, Step 2 in Section~\ref{subsec:cefadec}. 
%Let us use $\ostrichrecl_{\rm -SIMP}$ to denote $\ostrichrecl$ with the size-reduction techniques removed. 
We compare $\ostrichrecl$ with its variant  ${\rm +NUXMV}$. ${\rm +NUXMV}$  is obtained from $\ostrichrecl$ by using the nuXmv model checker to solve the nonemptiness of $\cefadec$.

@@

Since the non-register-representable regexes collected from real world are few (5\% of RegCoL and AutomatArk benchmark suites) and simple, there seems no much difference between the performance of $\ostrichrecl$ and its variants ${\rm -NEST}$, ${\rm -COMP}$, ${\rm -FIND}$ in the overall experiment results. Therefore, we consider the NestC benchmark suite, which contains more complex non-register-representable regexes, to justify the optimizations for non-register-representable regexes. The NestC benchmark experiment results is shown in Table~\ref{tab:results_heuristic_complex}. We can see that $\ostrichrecl$ solves 82/339/326 more instances and is 1.02/1.92/1.61 times faster than ${\rm -NEST}$/${\rm -COMP}$/${\rm -FIND}$ respectively. From the results, we can conclude that all the optimizations sharply improve the performance of $\ostrichrecl$ on the non-register-representable regexes.

\begin{table}
  \import{tables}{table_heuristics_complex.tex}
  \caption{Empirical justification of the optimizations on NestC benchmarks}\label{tab:results_heuristic_complex}
\end{table}
